---
title: "PU5558-PU5567-computer-lab-report"
author: "Romina-Degan"
output: pdf_document
---

## Introduction 

The following report is based on Provisional Patient Reported Outcome Measures (PROMS) specifically on the outcome of Hip and Knee Replacement procedures completed between April 2020 to March 2021,
provided by NHS England. For this report the focus will be on the outcome measures of knee replacement surgery. Specific outcome measures pertaining to the 
Quality Of Life (QOL) to the patient have been recorded through a visual analouge scale designed to measure the self-reported status of the patients quality of life pre/post surgery. These values 
gathered from the survery will be used to inform the models.  


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Load necessary packages

```{r load libraries, message=FALSE}
library(tidyverse)
library(tidymodels)
library(corrplot)
library(psych)
library(ggplot2)
library(rmarkdown)
library(reshape2)
library(vip)
library(Hmisc)
library(VIM)
library(caret)
library(flextable)
library(gtsummary)
library(randomForest)
```

## Load chosen dataset



```{r load data}
kneeData <- read_csv("Knee_Replacement_CCG 2021.csv")
kneeData <- kneeData %>%
    select(-Procedure, -Year, -"Revision Flag",  -"Provider Code") %>%
    drop_na() %>%
    unique()
```

\newpage
```{r display summary of data}
glimpse(kneeData)
# glimpseData <- kneeData[, c("Age Band", "Gender", "Pre-Op Q Assisted", "Heart Disease", "Arthritis",
#      "Post-Op Q Bleeding", "Pre-Op Q EQ5D Index Profile", "Post-Op Q EQ5D Index Profile",
#      "Pre-Op Q EQ5D Index", "Post-Op Q EQ5D Index")]
# kable(glimpseData, format = "latex", booktabs = TRUE, longtable = TRUE, 
#     caption="Rate of Change within an Individual Health Board Against its Moving Average") %>%
#     kable_styling(latex_options = c("striped", "hold_position")) %>%
#     row_spec(0, bold = TRUE) %>%
#     column_spec(1, "2cm") %>%
#     row_spec(15, bold = TRUE) %>%
#     kableExtra::landscape()

# set_flextable_defaults(
#     font.font.family = "Arial", font.size = 10,
#     border.color = "gray", big.mark = ""
# )


# ft <- flextable(head(kneeData), col_keys = c(
#     "Age Band", "Gender", "Pre-Op Q Assisted", "Heart Disease", "Arthritis",
#     "Post-Op Q Bleeding", "Pre-Op Q EQ5D Index Profile", "Post-Op Q EQ5D Index Profile",
#     "Pre-Op Q EQ5D Index", "Post-Op Q EQ5D Index"
# )) |> bold(part = "header")
# set_table_properties(ft, width=.5, layout = "autofit")
```
## Dataset description

This dataset uses data from NHS England regarding PROMS for knee replacement surgery. The way in which this data has been gathered is to collect self-reported symptoms/QOL metrics 
through two standard procedures. One survey is the EQ-5DTM Index designed to evaluate the QOL in terms of of doing daily tasks,such as(cite NHS PROMS): 
* Ability to pursue usual activites 
* Current experience of anxiety and/or depression, if any
* Current experience of pain and discomfort if any 
* Mobility
* Ability to wash and dress themselves
The closer the EQ5 value is to 1, the better the patients QOL. 

While the EQ5 focuses on day to day tasks the EQ-VAS focuses on pre/post operative outcomes through a visual thermomemter scale rannging from 0 to 100. This scale is designed to aid 
patients in describing how they feel post operation about the results and if their problems pre operation have been addressed with the surgery. 

Visually inspecting the database we see that Age Band and Gender have many missing values. Alongside this, there are many missing values for co-morbidites. As instructed all missing values for 
co-morbidites denotes as 9 will be taken as a no value. However, since Age Band and Gender typically are important factors (especially as confounders), these values will be imputed. Since there are 
a possible 77 variables to provide summary statstics on, only specific columns relevant to the following questions below will be considered.  
```{r display summary statstics}


kneeData |> tbl_summary(include=c("Age Band","Gender","Knee Replacement Post-Op Q Pain","Post-Op Q EQ5D Index","Pre-Op Q Previous Surgery" ))

```
## Suitable machine learning algorithm for three questions:

***1. Before the operation, can we estimate the post-operative EQ5D index for a patient?***
The EQ5 index is provided as a float value between 0 and 1. Due to its continous nature a suitable supervised model to answer this question would be a linear regression model. 
Since we wish to predict a continous variable against possible predictor variables pre/post surgery, we would have to find correlations between an increasing EQ5 index value 
and variables in the dataset. There are 2 possible approaches to answering this question: a simple linear regression model and a random forest for regression. Both models wont be made, 
however assumptions on the dataset will guide the choice of model; if the dataset is found to have a non-linear relationship or standard test to check for normality does not hold true, a 
random forest will be used. A linear regression model is simpler to run and provides better interpratibility of the model, however if assumptions do not hold, a random forest for regression 
must be used. This question will be used to complete the report.

***2. Before the operation, can we predict how much pain a patient will have after the operation?***
Since pain is measured as ordinal categories, a model designed for classification would fit this specification. Referencing (mention table figure) we see that 25% of data points are missing values 
for the questions regarding pain. If we were to use an unsupervised model it may not capture the true relationship between predictors and the outcome variable due to the vast number of unlabelled data points for a siginificant predictor. To 
tackle this issue of missing data, a semi-supervised model such as a semi-supervised support vector machine can overcome the missing data barrier. The semi-supervised SVM would use 
a mixture of labelled an unlabelled data to create descion boundaries for the different pain categories, and predict the positioning of a data point to its relative class given the presence 
of certain predictor variables. The level of computation a semi-supervised SVM has may be determinal to the models overall performance, but coupled with the fact that there are many other missing 
values in other questions regarding the QOL status of the patient, imputation on specific question would be difficult to complete without clinical knowledge to inform the possible current state of the patient. Therefore, 
the increased computational power for the model may be benefical when trying to address the gaps in data. 

***3. Before the operation, can we calculate how many patients have had previous surgery?***



Here we can use supervised/semi-supervised learning methods to cualcyate the numerical value of the
patiens that have had the previous surgery. The results is binary wether they have or have not had it 





## Model building to answer chosen question

# 1.	Data splitting 

Before the data is split into relevant train/test splits. The issue of Age Band and Gender both missing 35% of their data needs to be a handled. The following imputation is simple and doesnt 
use external questions or factors to inform descion making. Age Band is recoded into group numbers and the median of the group numbers is taken. Then the median value is then imputed to the missing data. 

Imputation of the Gender column is slightly more complex, with the method being a k Nearest Neighbour (kNN) approach. This approach tries to maintain the randomness of the gender data points, 
(reference table) looking at the table there are 6% more females that get the surgery than males. Therefore using a simple 50-50 or a median split would lose the close difference in value between 
the different genders. The KNN maintian thre nuisance in this split.
The following table summaries the changes in the data values after imputation, the changes are as follows
* Age Band - Median: Category 3, increase of 35% or 1,603 data points
* Gender - Male: Increase of 693 cases (+26%), Female: Increase of 910 cases (+20%)

Replacing missing values was only complete for age and gender since these values can be more easily accounted for in comparison to missing questionare values. Replacing missing questionare values
may have serious consequences to the predictions of the model, creating false equivalences. With certain predictors given more importance due to the inputted data. Removing these data values was not an option, 
since we would be losing up to 80% of the questionare data with some questions (hilight table). To ensure that the integrity of the the patient reported outcomes remain the same, more emphasis 
will be placed on the models to handle missing data such as a random forest (RF) model for regression. RF models dont need imputation of missing values and they dont suffer 
from underpreforming from the lack of labelled data points. This is due to their automatic handiling of missing data through surrogate splits. The randomForest package in R states in their documentation 

```{r imputation of data}
#Recode age bands into numeric groupings for an easier interpreation of the median
kneeData["Age Band"][kneeData["Age Band"] == "50 to 59"] <- "1"
kneeData["Age Band"][kneeData["Age Band"] == "60 to 69"] <- "2"
kneeData["Age Band"][kneeData["Age Band"] == "70 to 79"] <- "3"
kneeData["Age Band"][kneeData["Age Band"] == "80 to 89"] <- "4"
kneeData$"Age Band" <- as.numeric(as.character(kneeData$"Age Band"))

#Impute median values into the missing age data
kneeData["Age Band"][is.na(kneeData["Age Band"])] <- median(kneeData$"Age Band", na.rm = TRUE)
kneeData[kneeData == "*"] <- NA

#Call a KNN function designed to split the 
imputedGender <- kNN(kneeData, k = 1)
kneeData$Gender <- imputedGender$Gender
kneeData$Gender <- as.numeric(kneeData$Gender)

kneeData |> tbl_summary(include=c("Age Band","Gender" ))
kneeSplit <- kneeData %>%
    initial_split(prop = 0.7, strata = "Post-Op Q EQ5D Index")
    kneeTrain <- training(kneeSplit)
    kneeTest <- testing(kneeSplit)  
```

2.	Selection and preprocessing of predictors 

```{r correlation analysis}

kneeCorr <- cor(kneeTrain)
corrplot(kneeCorr, method = "color", type = "lower", tl.col = "black", tl.srt = 50, number.cex = 0.5, tl.cex = 0.3)


kneeCorrPostOp <- t(kneeCorr[, c(46), drop = FALSE])
corrplot(kneeCorrPostOp,method = "number", tl.col = "black", tl.srt = 50,tl.cex = 0.3, cex.axis=0.3, number.cex = 0.2)
for (i in seq_along(kneeTrain)) {
  cat("Index:", i, "- Name:", colnames(kneeTrain)[i], "\n")
}

```

The full correlation plot is displayed above but the values and their labels are difficult to read with the 77 variables that we have to consider. Extrcation of siginificant 
correlations in the main plot will be used to hilight the variables that will be used as predictors for the model. Correlation values were observed through the plots and 
all predictor values that were above 0.5 or below -0.5 were considered as significant predictors. This is puerly numerically based, but given clinical expertise there 
may be some predictors that may want to be included into the final model.  


```{r specifiy-recipie}
colnames(kneeTrain) <- make.names(colnames(kneeTrain))
colnames(kneeTest) <- make.names(colnames(kneeTest))

kneeModelRec <- kneeTrain %>%
    recipe(Post.Op.Q.EQ5D.Index ~
        Post.Op.Q.Mobility +
        Post.Op.Q.Self.Care +
        Post.Op.Q.Discomfort + 
        Post.Op.Q.EQ5D.Index.Profile +
        Post.Op.Q.EQ.VAS +
        Knee.Replacement.Post.Op.Q.Pain + 
        Knee.Replacement.Post.Op.Q.Night.Pain +
        Knee.Replacement.Post.Op.Q.Washing +
        Knee.Replacement.Post.Op.Q.Transport +
        Knee.Replacement.Post.Op.Q.Walking +
        Knee.Replacement.Post.Op.Q.Standing +
        Knee.Replacement.Post.Op.Q.Limping +
        Knee.Replacement.Post.Op.Q.Work +
        Knee.Replacement.Post.Op.Q.Confidence +
        Knee.Replacement.Post.Op.Q.Shopping +
        Knee.Replacement.Post.Op.Q.Stairs +
        Knee.Replacement.Post.Op.Q.Score) %>%
    step_zv(all_predictors()) %>%
    step_corr(all_predictors())

```
3.	Model specification and training

```{r call model}
kneeRandForest <- rand_forest() %>%
    set_mode("regression") %>%
    set_engine("randomForest")

kneeWorkflow <- workflow() %>%
    add_recipe(kneeModelRec) %>%
    add_model(kneeRandForest)

kneeFitModel <- kneeWorkflow %>%
    fit(data = kneeTrain)
```

```{r identify important vraibles}
kneeFitModel %>%
    extract_fit_parsnip() %>%
    vip(num_features=4)
```

4.	Model evaluation 
Code has 

```{r}
predEQVal <- kneeFitModel %>%
    predict(new_data = kneeTest)

kneeResult <- kneeTest %>%
    bind_cols(predEQVal) %>%
    rename(predEQ = .pred)

metrics(kneeResult, truth = Post.Op.Q.EQ5D.Index, estimate = predEQ)
```

```{r}
kneeResult %>%
    ggplot(aes(x = Post.Op.Q.EQ5D.Index, y = predEQ)) +
    geom_abline(intercept = 0, slope = 1) +
    geom_point() +
    xlab("Actual Outcome Values")
    ylab("Predicted Outcome Values")

```
<!-- # ```{r cross fold k validation models}
# crossVal <- trainControl(method = "cv", number = 8)

# kFoldModel <- 

# ``` -->

## Limitations of machine learning model

