---
title: "PU5558-PU5567-computer-lab-report"
author: "Anonymous"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load necessary packages

```{r}
library(tidyverse)
library(tidymodels)
library(corrplot)
library(psych)
library(ggplot2)
library(reshape2)
library(vip)
library(Hmisc)
library(VIM)
```

## Load chosen dataset

```{r}
kneeData <- read_csv("Knee_Replacement_CCG 2021.csv")
kneeData <- kneeData %>%
    select(-Procedure, -Year, -"Revision Flag",  -"Provider Code") %>%
    drop_na() %>%
    unique()
glimpse(kneeData)
```

## Dataset description
```{r}
describe(kneeData, fast=TRUE)
```

The following dataset depicts 
## Suitable machine learning algorithm for three questions:

1. Before the operation, can we estimate the post-operative EQ5D index for a patient?
Given certain presence of certain predictors being present, this would fall into supervised learning there are no missing values so no semi supervised is needed
0.364, 0.691, can use simple linear regression then test how well the model preforms with different AIC values. This question will be used for this report  

2. Before the operation, can we predict how much pain a patient will have after the operation?
Since pain is measured in categories we would want a classifier to predict which index the post operative pain will be
in this case we would use an unsupervised algorithim, seems intuitive that you would use forests

0 = Severe
1 = Moderate
2 = Mild
3 = Very Mild
4 = None
9 = Missing

3. Before the operation, can we calculate how many patients have had previous surgery? 
Here we can use supervised/semi-supervised learning methods to cualcyate the numerical value of the
patiens that have had the previous surgery. The results is binary wether they have or have not had it 





## Model building to answer chosen question

1.	Data splitting

```{r}
kneeData["Age Band"][kneeData["Age Band"] == "50 to 59"] <- "1"
kneeData["Age Band"][kneeData["Age Band"] == "60 to 69"] <- "2"
kneeData["Age Band"][kneeData["Age Band"] == "70 to 79"] <- "3"
kneeData["Age Band"][kneeData["Age Band"] == "80 to 89"] <- "4"
kneeData$"Age Band" <- as.numeric(as.character(kneeData$"Age Band"))
kneeData["Age Band"][is.na(kneeData["Age Band"])] <- median(kneeData$"Age Band", na.rm = TRUE)
kneeData[kneeData=="*"] <- NA
imputedGender <- kNN(kneeData, k = 10)
kneeData$Gender <- imputedGender$Gender
kneeData$Gender <- as.numeric(kneeData$Gender)




kneeSplit <- kneeData %>%
    initial_split(prop = 0.8, strata = "Post-Op Q EQ5D Index")
    kneeTrain <- training(kneeSplit)
    kneeTest <- testing(kneeSplit)  
```

2.	Selection and preprocessing of predictors
A quick analysis of the data shows that there are serveral variables that have a high proportion of missing values. This is represented with the 
median value being the missing value "9" this level of missing data will have severe effects on the true relationship between the post-op pain and the 
independent variables. Deleting these values would not be suitable as they contain relevant information that could have affect post-op pain (reference how co-morbidites increase
complications/pain managment post-op). As to the specifications of this assignment these values will be recoded so the value "9" represents "No".

Categorical values in the Age Band need to be recoded into dummy variables for regression analysis, the following encodings have been used in replacement
of the original categories

1- 50 to 59 "\n"
2- 60-69 
3- 70-79
4- 80-89 

Both Gender and Age Band have a considerable percentage of missing values equating to almost half the dataset, these will be addressed with different methods. 
Age Band will use imputation in the form of imputating the median of the categories from the encodings (1-4). Gender will use KNN imputation to maintain some randomness to the distribution. 

```{r}

kneeCorr <- cor(kneeTrain)
corrplot(kneeCorr, method = "color", type = "lower", tl.col = "black", tl.srt = 50, number.cex = 0.5, tl.cex = 0.3)


kneeCorrPostOp <- t(kneeCorr[, c(46), drop = FALSE])
corrplot(kneeCorrPostOp,method = "number", tl.col = "black", tl.srt = 50,tl.cex = 0.3, cex.axis=0.3, number.cex = 0.2)
for (i in seq_along(kneeTrain)) {
  cat("Index:", i, "- Name:", colnames(kneeTrain)[i], "\n")
}

```

The full correlation plot is displayed above but the values and their labels are difficult to read with the 77 variables that we have to consider. Extrcation of siginificant 
correlations in the main plot will be used to hilight the variables that will be used as predictors for the model. Correlation values were observed through the plots and 
all predictor values that were above 0.5 or below -0.5 were considered as significant predictors. This is puerly numerically based, but given clinical expertise there 
may be some predictors that may want to be included into the final model.  


```{r specifiy-recipie}
colnames(kneeTrain) <- make.names(colnames(kneeTrain))
print(colnames(kneeTrain))

kneeModelRec <- kneeTrain %>%
    recipe(Post.Op.Q.EQ5D.Index ~
        Post.Op.Q.Mobility +
        Post.Op.Q.Self.Care +
        Post.Op.Q.Discomfort + 
        Post.Op.Q.EQ5D.Index.Profile +
        Post.Op.Q.EQ.VAS +
        Knee.Replacement.Post.Op.Q.Pain + 
        Knee.Replacement.Post.Op.Q.Night.Pain +
        Knee.Replacement.Post.Op.Q.Washing +
        Knee.Replacement.Post.Op.Q.Transport +
        Knee.Replacement.Post.Op.Q.Walking +
        Knee.Replacement.Post.Op.Q.Standing +
        Knee.Replacement.Post.Op.Q.Limping +
        Knee.Replacement.Post.Op.Q.Work +
        Knee.Replacement.Post.Op.Q.Confidence +
        Knee.Replacement.Post.Op.Q.Shopping +
        Knee.Replacement.Post.Op.Q.Stairs +
        Knee.Replacement.Post.Op.Q.Score) %>%
    step_zv(all_predictors()) %>%
    step_corr(all_predictors())

```
3.	Model specification and training

```{r}
linearModel <- linear_reg()

kneeLM <- workflow() %>%
    add_recipe(kneeModelRec) %>%
    add_model(linearModel)

kneeModelFit <- kneeLM %>%
    fit(data = kneeTrain)

fittedKneeModel <- extract_fit_engine(kneeModelFit)
plot(fittedKneeModel, 1)
plot(fittedKneeModel, 2)
plot(fittedKneeModel, 3)

kneeModelFit %>% extract_fit_parsnip() %>% vip(num_features=8)
```

4.	Model evaluation

```{r}
predictedEQ <- kneeModelFit %>%
    predict(new_data = kneeTrain)

resultsTrain <- kneeTrain %>%
    bind_cols(predictedEQ) %>%
    rename(kneeEQPred= .pred)
metrics(resultsTrain, truth=Post.Op.Q.EQ5D.Index, estimate= kneeEQPred)
```

## Limitations of machine learning model

